\chapter{Boxes}
\section{Boxing examples}

The numbering of the various boxes (such as Definitions, Theorems, Exercises, etc.) is automatically reset at the beginning of each chapter. This means that every time a new chapter starts, the box counter resets to 1, both for the boxes in Italian and those in English. It is important to note that the numbering of the boxes is independent between the two languages: if the command is used in Italian, the box will be numbered according to the Italian numbering system, while using the command in English will generate a box with an independent English numbering. As a result, the numbering between the Italian and English boxes is not synchronized. It is recommended to pay attention and maintain consistency in language usage to avoid confusion and ensure the numbering remains correct within each chapter.

\begin{definizione}{Un esempio di definizione}
  Questo è un esempio di una casella colorata con il titolo "Definizione" in italiano.
\end{definizione}

\begin{definition}{An example of a definition}
  This is an example of a colored box with the title "Definition" in English.
\end{definition}

\begin{teorema}{Un esempio di teorema}
  Questo è un esempio di una casella colorata con il titolo "Teorema" in italiano.
\end{teorema}

\begin{theorem}{An example of a theorem}
  This is an example of a colored box with the title "Theorem" in English.
\end{theorem}

\begin{corollario}{Un esempio di corollario}
  Questo è un esempio di una casella colorata con il titolo "Corollario" in italiano.
\end{corollario}

\begin{corollary}{An example of a corollary}
  This is an example of a colored box with the title "Corollary" in English.
\end{corollary}

\begin{osservazione}{Un esempio di osservazione}
  Questo è un esempio di una casella colorata con il titolo "Osservazione" in italiano.
\end{osservazione}

\begin{observation}{An example of an observation}
  This is an example of a colored box with the title "Observation" in English.
\end{observation}

\begin{esercizio}{Un esempio di esercizio}
  Questo è un esempio di una casella colorata con il titolo "Esercizio" in italiano.
\end{esercizio}

\begin{exercise}{An example of an exercise}
  This is an example of a colored box with the title "Exercise" in English.
\end{exercise}
